```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
There are 3-4 packages you will need to install for today's practical: `install.packages(c("xgboost", "eegkit", "forecast", "tseries", "caret"))` apart from that everything else should already be available on your system. 

If you are using a newer Mac you may have to also install [quartz](https://www.xquartz.org/) to have everything work (do this if you see errors about `X11` during install/execution).

I will endeavour to use explicit imports to make it clear where functions are coming from (functions without `library_name::` are part of base R or a function we've defined in this notebook).

```{r libraries, echo=FALSE}
# Using the same library we used earlier in the course for tabular data because we know it works!
library(xgboost)

# EEG manipulation library in R (although very limited compared to signal processing libraries available in other languages, matlab might actually still be a leader in this specific area)
library(eegkit)

# some time series functions (that we only skim the depths of)
library(forecast)
library(tseries)
library(caret)

# just tidyverse libraries that should already be installed
library(dplyr)
library(reshape2)
library(purrr)
library(ggplot2)
```

## EEG Eye Detection Data

One of the most common types of medical sensor data (and one that we talked about during the lecture) are Electroencephalograms (EEGs).  
These measure mesoscale electrical signals (measured in microvolts) within the brain, which are indicative of a region of neuronal activity.
Typically, EEGs involve an array of sensors (aka channels) placed on the scalp with a high degree of covariance between sensors.

As EEG data can be very large and unwieldy, we are going to use a relatively small/simple dataset today from [this paper](http://ehrai.com/su/pdf/aihls2013.pdf).

This dataset is a 117 second continuous EEG measurement collected from a single person with a device called a "Emotiv EEG Neuroheadset".
In combination with the EEG data collection, a camera was used to record whether person being recorded had their eyes open or closed. 
This was eye status was then manually annotated onto the EEG data with `1` indicated the eyes being closed and `0` the eyes being open.
Measures microvoltages are listed in chronological order with the first measured value at the top of the dataframe.

Let's parse the data directly from the `h2o` library's (which we aren't actually using directly) test data S3 bucket:

```{r parse_data}
eeg_url <- "https://h2o-public-test-data.s3.amazonaws.com/smalldata/eeg/eeg_eyestate_splits.csv"
eeg_data <- read.csv(eeg_url)

# add timestamp
Fs <- 117 / nrow(eeg_data)
eeg_data <- transform(eeg_data, ds = seq(0, 116.99999, by = Fs), eyeDetection = as.factor(eyeDetection))
print(table(eeg_data$eyeDetection))

# split dataset into train, validate, test
eeg_train <- subset(eeg_data, split == 'train', select = -split)
print(table(eeg_train$eyeDetection))

eeg_validate <- subset(eeg_data, split == 'valid', select = -split)
eeg_test <- subset(eeg_data, split == 'test', select = -split)
```

**0** Knowing the `eeg_data` contains 117 seconds of data, inspect the `eeg_data` dataframe and the code above to and determine how many samples per second were taken?

```{r}

total_rows <- nrow(eeg_data)


total_seconds <- 117

samples_per_second <- total_rows / total_seconds


cat("Total number of rows (samples):", total_rows, "\n")
cat("Number of samples per second:", samples_per_second, "\n")

```






**1** How many EEG electrodes/sensors were used?
```{r}

column_names <- colnames(eeg_data)


sensor_columns <- setdiff(column_names, c("eyeDetection", "split", "ds"))


num_sensors <- length(sensor_columns)


cat("Number of EEG sensors used:", num_sensors, "\n")

```


### Exploratory Data Analysis

Now that we have the dataset and some basic parameters let's begin with the ever important/relevant exploratory data analysis.

First we should check there is no missing data!
```{r check_na}
sum(is.na(eeg_data))
```

Great, now we can start generating some plots to look at this data within the time-domain.

First we use `reshape2::melt()` to transform the `eeg_data` dataset from a wide format to a long format expected by `ggplot2`.

Specifically, this converts from "wide" where each electrode has its own column, to a "long" format, where each observation has its own row. 
This format is often more convenient for data analysis and visualization, especially when dealing with repeated measurements or time-series data.

We then use `ggplot2` to create a line plot of electrode intensities per sampling time, with the lines coloured by electrode, and the eye status annotated using dark grey blocks.

```{r plot_data}
melt <- reshape2::melt(eeg_data %>% dplyr::select(-split), id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")


ggplot2::ggplot(melt, ggplot2::aes(x=ds, y=microvolts, color=Electrode)) + 
  ggplot2::geom_line() + 
  ggplot2::ylim(3500,5000) + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(melt, eyeDetection==0), alpha=0.005)
```

**2** Do you see any obvious patterns between eyes being open (dark grey blocks in the plot) and the EEG intensities?

The dark grey blocks indicate periods when the eyes are open.
. During the eyes-open periods (dark grey blocks), the EEG signal amplitudes are generally lower compared to when the eyes are closed.
. The EEG signals are smoother and have fewer sharp peaks when the eyes are open.
. Some spikes, likely due to muscle movements or blinking, are more noticeable when the eyes are closed.

**3** Similarly, based on the distribution of eye open/close state over time to anticipate any temporal correlation between these states?

So the patterns we can identify are :
- If the dark grey (eyes open) and light grey (eyes closed) blocks alternate consistently, it indicates a predictable cycle of eye state changes.

- If the grey blocks are grouped together for longer stretches, it means the eyes stay open or closed for longer periods before switching states.

- If there are many short transitions between dark and light grey, it suggests frequent changes in eye states, possibly due to activities causing rapid blinking.

Let's see if we can directly look at the distribution of EEG intensities and see how they related to eye status.


As there are a few extreme outliers in voltage we will use the `dplyr::filter` function to remove values outwith of 3750 to 50003. The function uses the `%in%` operator to check if each value of microvolts is within that range. The function also uses the `dplyr::mutate()` to change the type of the variable eyeDetection from numeric to a factor (R's categorical variable type).

```{r compare_distrib}
melt_train <- reshape2::melt(eeg_train, id.vars=c("eyeDetection", "ds"), variable.name = "Electrode", value.name = "microvolts")

# filter huge outliers in voltage
filt_melt_train <- dplyr::filter(melt_train, microvolts %in% (3750:5000)) %>% dplyr::mutate(eyeDetection=as.factor(eyeDetection))

ggplot2::ggplot(filt_melt_train, ggplot2::aes(y=Electrode, x=microvolts, fill=eyeDetection)) + ggplot2::geom_boxplot()
```



Plots are great but sometimes so it is also useful to directly look at the summary statistics and how they related to eye status.
We will do this by grouping the data based on eye status and electrode before calculating the statistics using the convenient `dplyr::summarise` function.

```{r compare_summary_stats}
filt_melt_train %>% dplyr::group_by(eyeDetection, Electrode) %>% 
    dplyr::summarise(mean = mean(microvolts), median=median(microvolts), sd=sd(microvolts)) %>% 
    dplyr::arrange(Electrode)
```




**4** Based on these analyses are any electrodes consistently more intense or varied when eyes are open?

Electrodes `F8` and `AF4` exhibit higher intensity and variation when eyes are open, as indicated by higher means and standard deviations. Specifically, `F8` has a mean of 4615.185 and a standard deviation of 35.10948 with eyes open, compared to 4595.735 and 32.4387 with eyes closed. `AF4` shows a significant increase in variation with eyes open (46.58774) compared to eyes closed (36.03116). Other electrodes like `F7` and `F3` show less variability with eyes open. Overall, intensity and variation differences are electrode-specific.

#### Time-Related Trends

As it looks like there may be a temporal pattern in the data we should investigate how it changes over time.  

First we will do a statistical test for stationarity:

```{r convert_to_tseries}
apply(eeg_train, 2, tseries::adf.test)
```


**5** What is stationarity?


Stationarity is an important concept in time series analysis. A time series is considered stationary if its key characteristics—mean, variance, and the way values relate to each other over time—remain consistent throughout. In simpler terms, a stationary time series looks flat, without any upward or downward trends, and it doesn't show seasonal patterns or changes in its variability. 

**6** Why are we interested in stationarity? What do the results of these tests tell us? (ignoring the lack of multiple comparison correction...)

We are interested in stationarity because stationary series are easier to handle mathematically and ensure consistent relationships, as the correlations between observations remain constant over time. This consistency is crucial because many statistical tests and models assume or require stationarity to yield valid outcomes.

The ADF test outcomes bascially shows that the majority of the EEG sensor data are stationary, rendering them fit for subsequent time series analysis. Conversely, the ds variable is non-stationary and may need to be transformed, perhaps by differencing, to achieve stationarity before further analysis can be conducted.


Then we may want to visually explore patterns of autocorrelation (previous values predict future ones) and cross-correlation (correlation across channels over time) using `forecast::ggAcf` function.

The ACF plot displays the cross- and auto-correlation values for different lags (i.e., time delayed versions of each electrode's voltage timeseries) in the dataset. 
It helps identify any significant correlations between channels and observations at different time points. 
Positive autocorrelation indicates that the increase in voltage observed in a given time-interval leads to a proportionate increase in the lagged time interval as well.
Negative autocorrelation indicates the opposite!


```{r correlation}
forecast::ggAcf(eeg_train %>% dplyr::select(-ds))
```





**7** Do any fields show signs of strong autocorrelation (diagonal plots)? Do any pairs of fields show signs of cross-correlation? Provide examples.

The diagonal plot of F7 shows strong spikes at various lags, indicating significant autocorrelation. Additionally, fields such as F3, FC5, T7, P7, T8, FC6, F4, F8, AF4, and eyeDetection also exhibit strong signs of autocorrelation with significant spikes.

The off-diagonal plot between F7 and F3 shows consistent spikes and patterns, indicating cross-correlation. Additionally, pairs such as F7-F3, F3-FC5, T7-P7, T8-FC6, and F4-F8 exhibit noticeable patterns in the off-diagonal plots, further suggesting cross-correlation.
#### Frequency-Space 

We can also explore the data in frequency space by using a Fast Fourier Transform.  
After the FFT we can summarise the distributions of frequencies by their density across the power spectrum.
This will let us see if there any obvious patterns related to eye status in the overall frequency distributions.

```{r fft_open}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 0) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Open")
```

```{r fft_closed}
eegkit::eegpsd(eeg_train %>% dplyr::filter(eyeDetection == 1) %>% dplyr::select(-eyeDetection, -ds), Fs = Fs, xlab="Eye Closed")
```




**8** Do you see any differences between the power spectral densities for the two eye states? If so, describe them.

The "Eye Open" state exhibits generally lower power levels across most channels, as indicated by the predominance of blue and cyan colors, which represent lower power. In this state, higher power levels (orange) are mostly concentrated at the edges of the plot, indicating elevated power at specific frequencies or time points. Conversely, the "Eye Closed" state shows higher power levels overall, with more widespread regions of elevated power (orange) across multiple channels, suggesting an increase in power across a broader range of frequencies.

These variations indicate changes in brain activity between the two eye states. This observation aligns with typical findings in EEG studies, where closing the eyes is often linked to an increase in alpha wave activity.

#### Independent Component Analysis

We may also wish to explore whether there are multiple sources of neuronal activity being picked up by the sensors.  
This can be achieved using a process known as independent component analysis (ICA) which decorrelates the channels and identifies the primary sources of signal within the decorrelated matrix.

```{r ica, warning=FALSE}
ica <- eegkit::eegica(eeg_train %>% dplyr::select(-eyeDetection, -ds), nc=3, method='fast', type='time')
mix <- dplyr::as_tibble(ica$M)
mix$eyeDetection <- eeg_train$eyeDetection
mix$ds <- eeg_train$ds

mix_melt <- reshape2::melt(mix, id.vars=c("eyeDetection", "ds"), variable.name = "Independent Component", value.name = "M")


ggplot2::ggplot(mix_melt, ggplot2::aes(x=ds, y=M, color=`Independent Component`)) + 
  ggplot2::geom_line() + 
  ggplot2::geom_vline(ggplot2::aes(xintercept=ds), data=dplyr::filter(mix_melt, eyeDetection==1), alpha=0.005) +
  ggplot2::scale_y_log10()
```



**9** Does this suggest eye opening relates to an independent component of activity across the electrodes?

The graph indicates a significant independent component (V1) of activity that likely relates to eye opening across the electrodes. The noticeable fluctuations and spikes in V1 suggest it captures events associated with changes in eye state. The presence of gray bands (V2 and V3) further supports the analysis of multiple components, hinting at a correlation between V1 activity and eye movements or transitions.

The gray bands probably signify other independent components (V2 and V3). While their activities are not explicitly shown, their presence indicates that multiple components are being analyzed concurrently.


### Eye Opening Prediction

Now that we've explored the data let's use a simple model to see how well we can predict eye status from the EEGs:

```{r xgboost}
# Convert the training and validation datasets to matrices
eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.numeric(eeg_train$eyeDetection) -1

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.numeric(eeg_validate$eyeDetection) -1

# Build the xgboost model
model <- xgboost(data = eeg_train_matrix, 
                 label = eeg_train_labels,
                 nrounds = 100,
                 max_depth = 4,
                 eta = 0.1,
                 objective = "binary:logistic")

print(model)
```



**10** Using the `caret` library (or any other library/model type you want such as a naive Bayes) fit another model to predict eye opening.

```{r model2}


eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.factor(eeg_train$eyeDetection)

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.factor(eeg_validate$eyeDetection)


eeg_train_df <- data.frame(eeg_train_matrix)
eeg_train_df$eyeDetection <- eeg_train_labels

eeg_validate_df <- data.frame(eeg_validate_matrix)
eeg_validate_df$eyeDetection <- eeg_validate_labels


set.seed(123) 
model <- train(eyeDetection ~ ., data = eeg_train_df, method = "rf", 
               trControl = trainControl(method = "cv", number = 10))


print(model)


predictions <- predict(model, newdata = eeg_validate_df)


confusionMatrix(predictions, eeg_validate_labels)


```


**11** Using the best performing of the two models (on the validation dataset) calculate and report the test performance (filling in the code below):

```{r test}




eeg_train_matrix <- as.matrix(dplyr::select(eeg_train, -eyeDetection, -ds))
eeg_train_labels <- as.numeric(eeg_train$eyeDetection) - 1

eeg_validate_matrix <- as.matrix(dplyr::select(eeg_validate, -eyeDetection, -ds))
eeg_validate_labels <- as.numeric(eeg_validate$eyeDetection) - 1


model <- xgboost(data = eeg_train_matrix,
                 label = eeg_train_labels,
                 nrounds = 100,
                 max_depth = 4,
                 eta = 0.1,
                 objective = "binary:logistic")


pred_val <- predict(model, eeg_validate_matrix)
pred_val <- ifelse(pred_val > 0.5, 1, 0)


accuracy_val <- mean(pred_val == eeg_validate_labels)
cat("Validation Accuracy:", accuracy_val, "\n")




eeg_test_matrix <- as.matrix(dplyr::select(eeg_test, -eyeDetection, -ds))
eeg_test_labels <- as.numeric(eeg_test$eyeDetection) - 1


pred_test <- predict(model, eeg_test_matrix)
pred_test <- ifelse(pred_test > 0.5, 1, 0)


accuracy_test <- mean(pred_test == eeg_test_labels)
cat("Test Accuracy:", accuracy_test, "\n")


```

**12** Describe 2 possible alternative modeling approaches for prediction of eye opening from EEGs we discussed in the lecture but haven't explored in this notebook.

RNNs is ideal for analyzing EEG signals because it can capture the sequential nature of the data. EEG recordings consist of time-series measurements where the order and timing are critical.

CNNs, typically applied in image processing, can be tailored for 1D signals such as EEGs. By employing filters across the input sequence, CNNs excel at identifying intricate patterns, making them valuable for extracting both spatial and temporal characteristics from EEG data.

Also Differencing and AutoRegressive models can be alternative modeling approaches Differencing involves taking the difference between successive observations in a time series. and AR models are linear models that use previous observations (lags) of a time series to predict future values. 


**13** What are 2 R libraries you could use to implement these approaches? (note: you don't actually have to implement them though!)

Two R libraries that could be used to implement these approaches are Recurrent Neural Networks and Random Forests.

## Optional

**14** (Optional) As this is the last practical of the course - let me know how you would change future offerings of this course. This will not impact your marks!

- What worked and didn’t work for you (e.g., in terms of the practicals, tutorials, and lectures)?

- Was learning how to run the practicals on your own machines instead of a clean server that will disappear after the course worth the technical challenges?
 
- What would you add or remove from the course? 

- What was the main thing you will take away from this course?