```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(tidytext)
library(textstem)
library(clinspacy)
library(topicmodels)
library('reshape2')
library(stringr)

```

This practical is based on exploratory data analysis, named entity recognition, and topic modelling of unstructured medical note free-text data derived from electronic medical records (EMR).  Real EMR data is very difficult to access without a specific need/request so this data set is derived from [medical transcription](https://mtsamples.com/) data instead.  I'll also caveat that the options of natural language processing (NLP) in R are far inferior to those available in Python. 

First, install the packages in the setup block (`install.packages(c("readr", "dplyr", "tidyr", "ggplot2", "tidtext", "textstem", "clinspacy", "topicmodels", "reshape2"))`).

Note: To try and make it clearer which library certain functions are coming from clearer, I'll try to do explicit imports throughout this notebook.

## Data Parsing

After that we can grab the dataset directly from the `clinspacy` library.

```{r}
raw.data <- clinspacy::dataset_mtsamples()
dplyr::glimpse(raw.data)
```

There is no explanation or data dictionary with this dataset, which is a surprisingly common and frustrating turn of events!  

**1** Using the output of dplyr's `glimpse` command (or rstudio's data viewer by clicking on `raw.data` in the Environment pane) provide a description of what you think each variable in this dataset contains.

note_id is basically a unique identifier assigned to each note.
description gives a brief summary of the note's content.
medical_specialty classifies the note according to its relevant medical field.
sample_name provides a descriptive title for the note which describes it.
transcription includes the comprehensive text of the medical note.
keywords highlight the key terms associated with the note's content.


Let's see how many different medical specialties are featured in these notes: 
```{r}

raw.data %>% dplyr::select(medical_specialty) %>% dplyr::n_distinct()
```

So, how many transcripts are there from each specialty:


```{r}
ggplot2::ggplot(raw.data, ggplot2::aes(y=medical_specialty)) + ggplot2::geom_bar() + labs(x="Document Count", y="Medical Speciality")
```

Let's make our life easier and filter down to 3 specialties: a diagonstic/lab, a medical, and a surgical specialty

```{r} 
filtered.data <- raw.data %>% dplyr::filter(medical_specialty %in% c("Orthopedic", "Radiology", "Surgery")) 
```

## Text Processing

Let's now apply our standard pre-processing to the transcripts from these specialties.  
We are going to use the `tidytext` package to tokenise the transcript free-text.  
Let's remove stop words first. e.g., "the", "of", "to", and so forth. These are known as stop words and we can remove them relative easily using a list from  `tidytext::stop_words` and `dplyr::anti_join()`

```{r}

analysis.data <- filtered.data %>%
  unnest_tokens(word, transcription) %>%
  mutate(word = str_replace_all(word, "[^[:alnum:]]", "")) %>%
  filter(!str_detect(word, "[0-9]")) %>%
  anti_join(stop_words) %>%
  group_by(note_id) %>%
  summarise(transcription = paste(word, collapse = " ")) %>%
  left_join(select(filtered.data, -transcription), by = "note_id")
```


Now let's tokenize the `transcription` to words (unigram) 
By default this tokenises to words but other options include characters, n-grams, sentences, lines, paragraphs, or separation around a regular expression.

```{r}
tokenized.data.unigram <- analysis.data %>% tidytext::unnest_tokens(word, transcription, to_lower=TRUE)
```

You can also do bi-grams
```{r}
tokenized.data <- analysis.data %>% tidytext::unnest_tokens(ngram, transcription, token = "ngrams", n=2, to_lower = TRUE)
```

How many stop words are there in `tidytext::stop_words` from each lexicon?
```{r}
tidytext::stop_words %>% dplyr::group_by(lexicon) %>% dplyr::distinct(word) %>% dplyr::summarise(n=dplyr::n())
```

**2** How many unique unigrams are there in the transcripts from each specialty:

```{r}
#Insert code here. 
tokenized.data.unigram %>% dplyr::group_by(medical_specialty) %>% dplyr::distinct(word) %>% dplyr::summarise(n=dplyr::n())
```

There are 
Let's plot some distribution of unigram tokens (words)


```{r}
word_counts <- tokenized.data.unigram %>%
    group_by(word) %>%
    summarise(count = n()) %>%
    ungroup() %>%
    arrange(desc(count))

count_distribution <- word_counts %>%
  group_by(count) %>%
  summarise(num_words = n()) %>%
  ungroup()
 
 ggplot2::ggplot(count_distribution, aes(x = count, y = num_words)) +
  geom_point() +
  labs(title = "Scatter Plot of Count Distribution",
       x = "Count of Unique Words",
       y = "Number of Words")
```


Let's plot some distribution of bigram tokens (words)


```{r}
word_counts <- tokenized.data %>%
    group_by(ngram) %>%
    summarise(count = n()) %>%
    ungroup() %>%
    arrange(desc(count))

count_distribution <- word_counts %>%
  group_by(count) %>%
  summarise(num_words = n()) %>%
  ungroup()
 
 ggplot2::ggplot(count_distribution, aes(x = count, y = num_words)) +
  geom_point() +
  labs(title = "Scatter Plot of Count Distribution",
       x = "Count of Unique Bigrams",
       y = "Number of Words")
```

**3** How many unique bi-grams are there in each category without stop words and numbers?
```{r}
#Insert code here. 
tokenized.data %>% dplyr::group_by(medical_specialty) %>% dplyr::distinct(ngram) %>% dplyr::summarise(n=dplyr::n())
```
Sometimes we are interested in tokenising/segmenting things other than words like whole sentences or paragraphs.  
**4** How many unique sentences are there in each category? Hint: use `?tidytext::unnest_tokens` to see the documentation for this function.

```{r}
tokenized.data.sentences <- analysis.data %>% tidytext::unnest_tokens(sentences, transcription, token = "sentences", to_lower = TRUE)
```
```{r}
#Insert code here. 
tokenized.data.sentences %>% dplyr::group_by(medical_specialty) %>% dplyr::distinct(sentences) %>% dplyr::summarise(n=dplyr::n())
```
Now that we've tokenized to words and removed stop words, we can find the most commonly word used within each category:

```{r}
tokenized.data %>%
  dplyr::group_by(medical_specialty) %>%
  dplyr::count(ngram, sort = TRUE) %>%
  dplyr::top_n(5)
```

We should lemmatize the tokenized words to prevent over counting of similar words before further analyses.  
Annoyingly, `tidytext` doesn't have a built-in lemmatizer.

**5** Do you think a general purpose lemmatizer will work well for medical data? Why might it not?

No, I don't think a general-purpose lemmatizer will work well for medical data. It might not perform effectively because it isn't trained on clinical language, which is not commonly used by the general public.They may use different terminology in different specialties and that the lemmatization may be better if it is built to that specific specialty and words that would be used within it.

Unfortunately, a specialised lemmatizer like in `clinspacy` is going to be very painful to install so we will just use a simple lemmatizer for now:

```{r}
lemmatized.data <- tokenized.data %>% dplyr::mutate(lemma=textstem::lemmatize_words(ngram))
```

We can now calculate the frequency of lemmas within each specialty and note.
```{r}
lemma.freq <- lemmatized.data %>% 
  dplyr::count(medical_specialty, lemma) %>%
  dplyr::group_by(medical_specialty) %>% 
  dplyr::mutate(proportion = n / sum(n)) %>%
  tidyr::pivot_wider(names_from = medical_specialty, values_from = proportion) %>%
  tidyr::pivot_longer(`Surgery`:`Radiology`,
               names_to = "medical_specialty", values_to = "proportion")
```

And plot the relative proportions 
```{r}

ggplot2::ggplot(lemma.freq, ggplot2::aes(x=proportion, 
                                         y=`Orthopedic`,
                                         color=abs(`Orthopedic` - proportion))) + 
  ggplot2::geom_abline(color="gray40", lty=2) +
  ggplot2::geom_jitter(alpha=0.1, size=2.5, width=0.3, height=0.3) +
  ggplot2::geom_text(ggplot2::aes(label=lemma), check_overlap=TRUE, vjust=1.5) +
  ggplot2::scale_x_log10(labels=scales::percent_format()) + 
  ggplot2::scale_y_log10(labels=scales::percent_format()) + 
  ggplot2::scale_color_gradient(limits=c(0, 0.001), low="darkslategray4", high="gray75") +
  ggplot2::facet_wrap(~medical_specialty, ncol = 2) +
  ggplot2::theme(legend.position="none") +
  ggplot2:: labs(y="Orthopedic", x = NULL)
```
**6** What does this plot tell you about the relative similarity of lemma frequencies between Surgery and Orthopedic and between radiology and Orthopedic? Based on what these specialties involve, is this what you would expect?
The plot demonstrates that lemma frequencies are more closely aligned between Radiology and Orthopedic than between Surgery and Orthopedic.
1. Orthopedic vs. Radiology 
So the left panel displays a comparison of lemma frequencies between Radiology and Orthopedic. Points near the diagonal line show that terms are used similarly in both specialties, with most terms appearing infrequently and clustered in the lower left corner. Some terms, such as "series images,"  and "acute fracture," are slightly more common. 

2. Surgery vs. Orthopedic 
So the right panel illustrates the comparison of lemma frequencies between Surgery and Orthopedic. The wider spread of points, with terms such as "carpal ligament" and "fracture site" being more common in Surgery, suggests a greater variation in term usage. 

**7** Modify the above plotting code to do a direct comparison of Surgery and Radiology (i.e., have Surgery or Radiology on the Y-axis and the other 2 specialties as the X facets)

```{r}
# Assuming 'lemmatized.data' contains the lemmatized ngrams
lemma.freq <- lemmatized.data %>%
  dplyr::count(medical_specialty, lemma) %>%
  dplyr::group_by(medical_specialty) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  tidyr::pivot_wider(
    names_from = medical_specialty, 
    values_from = proportion, 
    values_fill = list(proportion = 0)  # Filling missing values with 0
  )
 
 
# Check the names of the columns to ensure they are present
print(colnames(lemma.freq))
 
# Proceed with plotting if the columns are named correctly
ggplot(lemma.freq, aes(x = Surgery, y = Radiology, color = abs(Surgery - Radiology))) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray40") +
  geom_jitter(alpha = 0.1, size = 2.5, width = 0.3, height = 0.3) +
  geom_text(aes(label = lemma), check_overlap = TRUE, vjust = 1.5) +
  scale_x_log10(labels = scales::percent_format()) +
  scale_y_log10(labels = scales::percent_format()) +
  scale_color_gradient(limits = c(0, 0.001), low = "darkslategray4", high = "gray75") +
  labs(x = "Surgery Proportion", y = "Radiology Proportion") +
  theme(legend.position = "none")
 
```

### TF-IDF Normalisation

Maybe looking at lemmas across all notes in a specialty is misleading, what if we look at lemma frequencies across a specialty.

```{r}
lemma.counts <- lemmatized.data %>% dplyr::count(medical_specialty, lemma)
total.counts <- lemma.counts %>% 
                      dplyr::group_by(medical_specialty) %>% 
                      dplyr::summarise(total=sum(n))

all.counts <- dplyr::left_join(lemma.counts, total.counts)
```
Now we can calculate the term frequency / invariant document frequency (tf-idf):

```{r}
all.counts.tfidf <- tidytext::bind_tf_idf(all.counts, lemma, medical_specialty, n) 
```

We can then look at the top 10 lemma by tf-idf within each specialty:

```{r}
all.counts.tfidf %>% dplyr::group_by(medical_specialty) %>% dplyr::slice_max(order_by=tf_idf, n=10)
```
**8** Are there any lemmas that stand out in these lists? Why or why not?

The term "department anesthesia" stands out the most in these lists. Unlike other terms that are specifically related to orthopedic techniques, "department anesthesia" is a general term applicable to all surgical departments and procedures. It stands out due to its consistent presence and relevance across multiple entries, particularly in the Surgery specialty, highlighting its critical role in medical procedures and its contextual importance.

Lemmas That Stand Out:
"range motion"

TF-IDF Score: This term has a high relevance score because it is specifically used to assess joint functionality, which is crucial in orthopedic diagnostics and postoperative care.
Contextual Importance: It often appears in clinical notes about patient mobility issues, which are essential for diagnosing and planning treatment in musculoskeletal medicine.
"carpal ligament" and "transverse carpal"

Specificity: These terms are directly linked to carpal tunnel syndrome, a common orthopedic condition, making them significant in specialized texts.
Technical Relevance: They are used in detailed discussions about specific wrist areas, important for surgical planning and evaluating outcomes.
Lemmas That Do Not Stand Out and Why:
"steri strips" and "closed vicryl"

These terms are related to surgical procedures but are common across various types of surgeries, not just Orthopedics. Therefore, they might not stand out in a TF-IDF analysis that focuses on identifying unique terms in Orthopedic texts.
"proximal phalanx"

Although this term refers to a specific bone in the finger or toe and could be expected to stand out in Orthopedic discussions, its significance may be reduced if other specialties also frequently mention it in different contexts, affecting its TF-IDF score.

We can look at transcriptions in full using these lemmas to check how they are used with `stringr::str_detect`
```{r}
analysis.data %>% dplyr::select(medical_specialty, transcription) %>% dplyr::filter(stringr::str_detect(transcription, 'steri strips')) %>% dplyr::slice(1)
```

**9** Extract an example of one of the other "top lemmas" by modifying the above code
```{r}
analysis.data %>% dplyr::select(medical_specialty, transcription) %>% dplyr::filter(stringr::str_detect(transcription, 'dissection carried')) %>% dplyr::slice(1)
```

## Topic Modelling

In NLP, we often have collections of documents (in our case EMR transcriptions) that we’d like to divide into groups so that we can understand them separately. Topic modeling is a method for unsupervised classification of such documents, similar to clustering on numeric data.

Latent Dirichlet allocation (LDA) is a particularly popular method for fitting a topic model. It treats each document as a mixture of topics, and each topic as a mixture of words. This allows documents to “overlap” each other in terms of content, rather than being separated into discrete groups, in a way that mirrors typical use of natural language.


- Every document is a mixture of topics. We imagine that each document may contain words from several topics in particular proportions. For example, in a two-topic model we could say “Document 1 is 90% topic A and 10% topic B, while Document 2 is 30% topic A and 70% topic B.”


- Every topic is a mixture of words. For example, we could imagine a two-topic model of American news, with one topic for “politics” and one for “entertainment.” The most common words in the politics topic might be “President”, “Congress”, and “government”, while the entertainment topic may be made up of words such as “movies”, “television”, and “actor”. Importantly, words can be shared between topics; a word like “budget” might appear in both equally.

LDA is a mathematical method for estimating both of these at the same time: finding the mixture of words that is associated with each topic, while also determining the mixture of topics that describes each document. There are a number of existing implementations of this algorithm, and we’ll explore one of them in depth.

First lets calculate a term frequency matrix for each transcription:
```{r}

lemma.counts <- lemmatized.data %>% dplyr::count(note_id, lemma)
total.counts <- lemma.counts %>% 
                      dplyr::group_by(note_id) %>% 
                      dplyr::summarise(total=sum(n))

all.counts <- dplyr::left_join(lemma.counts, total.counts)

emr.dcm <- all.counts %>% tidytext::cast_dtm(note_id, lemma, n)
```

Then we can use LDA function to fit a 5 topic (`k=5`) LDA-model
```{r}
emr.lda <- topicmodels::LDA(emr.dcm, k=5, control=list(seed=42))
emr.topics <- tidytext::tidy(emr.lda, matrix='beta')
```

Then we can extract the top terms per assigned topic:
```{r}

top.terms <- emr.topics %>% dplyr::group_by(topic) %>% 
  dplyr::slice_max(beta, n=10) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(topic, -beta)


top.terms %>% 
  dplyr::mutate(term=tidytext::reorder_within(term, beta, topic)) %>% 
  ggplot2::ggplot(ggplot2::aes(beta, term, fill=factor(topic))) + 
    ggplot2::geom_col(show.legend=FALSE) + 
    ggplot2::facet_wrap(~ topic, scales='free')  +
    ggplot2::theme(axis.text.x = element_text(angle = 45,vjust = 1,hjust = 1)) +
    tidytext::scale_y_reordered()
```



Now we can ask how well do these assigned topics match up to the medical specialties from which each of these transcripts was derived.

```{r}
specialty_gamma <- tidytext::tidy(emr.lda, matrix='gamma')

# we need to join in the specialty from the note_id
note_id_specialty_mapping <- lemmatized.data %>%
  dplyr::mutate(document=as.character(note_id)) %>% 
  dplyr::select(document, medical_specialty) %>% 
  dplyr::distinct()

specialty_gamma <- dplyr::left_join(specialty_gamma, note_id_specialty_mapping)
```

```{r}

specialty_gamma %>%
  dplyr::mutate(medical_specialty = reorder(medical_specialty, gamma * topic)) %>%
  ggplot2::ggplot(ggplot2::aes(factor(topic), gamma)) +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~ medical_specialty) +
  ggplot2::labs(x = "topic", y = expression(gamma))
```

Interestingly, Surgery, Orthopedic, and Radiology assign mostly to a single topics. We'd possibly expect this from radiology due to referring to imaging for many different diagnoses/reasons. 
However, this may all just reflect we are using too few topics in our LDA to capture the range of possible assignments. 

**10** Repeat this with a 6 topic LDA, do the top terms from the 3 topic LDA still turn up? How do the specialties get split into sub-topics?

```{r}
emr.lda <- topicmodels::LDA(emr.dcm, k=6, control=list(seed=42))
emr.topics <- tidytext::tidy(emr.lda, matrix='beta')
```

Then we can extract the top terms per assigned topic:
```{r}

top.terms <- emr.topics %>% dplyr::group_by(topic) %>% 
  dplyr::slice_max(beta, n=10) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(topic, -beta)


top.terms %>% 
  dplyr::mutate(term=tidytext::reorder_within(term, beta, topic)) %>% 
  ggplot2::ggplot(ggplot2::aes(beta, term, fill=factor(topic))) + 
    ggplot2::geom_col(show.legend=FALSE) + 
    ggplot2::facet_wrap(~ topic, scales='free')  +
    ggplot2::theme(axis.text.x = element_text(angle = 45,vjust = 1,hjust = 1)) +
    tidytext::scale_y_reordered()
```

```{r}
specialty_gamma <- tidytext::tidy(emr.lda, matrix='gamma')

# we need to join in the specialty from the note_id
note_id_specialty_mapping <- lemmatized.data %>%
  dplyr::mutate(document=as.character(note_id)) %>% 
  dplyr::select(document, medical_specialty) %>% 
  dplyr::distinct()

specialty_gamma <- dplyr::left_join(specialty_gamma, note_id_specialty_mapping)
```

```{r}

specialty_gamma %>%
  dplyr::mutate(medical_specialty = reorder(medical_specialty, gamma * topic)) %>%
  ggplot2::ggplot(ggplot2::aes(factor(topic), gamma)) +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~ medical_specialty) +
  ggplot2::labs(x = "topic", y = expression(gamma))
```
## Credits

Examples draw heavily on material (and directly quotes/copies text) from Julia Slige's `tidytext` textbook.

